## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  a
} else {
  "it's more than 5"
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
  'small'
} else if (a > 5 & a <= 10) {
  'medium'
} else if (a > 10 | a == 7) {
  'large'
} else {
  'huge'
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a / 2 != 0) {
  if (a == 5) {
    'a is 5'
  }
}

```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  # paste is a concat function
  print(paste(value, 'is greater than', value -1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# change the 1st, 3rd, 6th elements to the value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10){
  # increase b by 2
  b <- b + 2
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}

```

An example:

```{r}
# start with a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820,
               21.173577, 21.048836, 18.769735)

# then we want a function that converts from C to F:
c_to_f <- function(dat){
  temp_fahr = dat*(9/5) + 32
  return(temp_fahr)
}

print(c_to_f(26))

```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow = 2)
test_mat

```

```{r}
# 1 is for rows, 2 is for columns
apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1 Convert a Number to a String!
    number_to_string <- function(n) {
     as.character(n)
    }

    n <- 123
    number_to_string(n)
    ```

    )

    ```{r}
    #2 Multiply
    mul <- function(a, b) {
      a * b # try to figure out why it doesn't work!
    }

    a <- 5
    b <- 10
    mul(a,b)
    ```

    ```{r}
    #3 Descending Order
    descending_order <- function(num){
      as.numeric(paste(sort(unlist(strsplit(as.character(num), NULL)), decreasing = TRUE), collapse = ""))
    }

    num <- 145263
    descending_order(num)
    ```

    ```{r}
    #4 Sum of positive
    positive_sum <- function(vec){
     sum(vec[vec > 0])
    }

    vec <- c(1,-4,7,12)
    positive_sum(vec)
    ```

    ```{r}
    #5 Highest and Lowest
    high_and_low <- function(numbers){
     nums <- as.integer(unlist(strsplit(numbers, " ")))
      paste(max(nums), min(nums)) 
    }

    numbers <-  "1 9 30 4 -5"
    high_and_low(numbers)
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fizzbuzz <- function(num){
      if (num %% 3 == 0 & num %% 5 == 0){
        "fizzbuzz"
      }else if (num %% 3 == 0){
        "fizz"
      }else if (num %% 5 == 0){
        "buzz"
      }else {
        "This number is neither fizz or buzz."
      }
    }

    num = 60
    fizzbuzz(num)
    num = 9
    fizzbuzz(num)
    num = 20
    fizzbuzz(num)
    num = 22
    fizzbuzz(num)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.
    marioStairs <- function(n){
      stairs_m <- matrix('', n, n)
      stairs_m[!upper.tri(stairs_m)[1:n,]] <- 'X'
      prmatrix(stairs_m, quote = FALSE, rowlab = rep('',n), collab = rep('', n))
    }

    marioStairs(4)
    ```
